\textls[-0]{\thetitle}

### 2023-03-13 metatheory-condensed.tex
Fix rule ref name

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT.
Together with the identity type $\TyId{\cDot}{\cDot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
Our enhanced W-type is decided by a W-type signature $\goodWSig{\Gm}{\wsig}{n}$ 
composed of a list (\ruleref{wsig/empty} and \ruleref{wsig/add}) of $n$ pairs of 
types $\{(\wsigproj{i}{1}{\wsig}, \wsigproj{i}{2}{\wsig})\}_i$,
modeling an inductive type of $n$ constructors. 
%
Rule~\ruleref{tm/wsup} states that the $i$-th pair $\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and 
$\goodType{\Gm,\wsigproj{i}{1}{\wsig}}{\wsigproj{i}{2}{\wsig}}{}$ models the $i$-th constructor---given 
two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm, \mathit{self} : \nsub{\wsigproj{i}{2}{\wsig}}{x}{t_1}}{t_2}{\El{\wcode{\wsig}}}$,
we can construct a term $\Wsup{i}{\wsig}{t_1}{\mathit{self}.t_2}$ of the W-type. $\wcode{\wsig}$ gives the code of the
W-type (\ruleref{tm/w}).

W-types are eliminated with the form $\Wrec{\wsig}{\lkg}{t}$, where $t$ is of a
W-type $\El{\wcode{\wsig}}$, and $\lkg$ is essentially an $n$-tuple of
case handlers for the $n$ constructors in~$\wsig$ (\ruleref{tm/wrec}).
Each case handler has a type of form $\CaseSig{A}{B}{T}$, where
$T$ is the motive of the recursion (\ruleref{tyeq/casety});
for simplicity, we model non-dependent motives for recursion.
The collection of case handlers~$\lkg$ encodes those defined and inherited
by an \lsti{FRecursion} command in our plugin.
We choose to type it with a linkage type $\TyLkg{\RecSig{\wsig}{T}}$
to avoid introducing non-dependent $n$-tuples, which linkages generalize.

@@

Rules \ruleref{tm/wsup} and \ruleref{tm/wrec} require access to~$\wsig$:
the W-type is exhaustively generated by its constructors, and
its elimination must exhaustively handle all the constructors in
its signature.

@@

Well-formedness judgments of \emph{linkage signatures} have the form
$\goodSig{\Gm}{\lsig}{n}$.
As rules \ruleref{lsig/empty} and \ruleref{lsig/add} show, a linkage
signature is a list of $n$~types.
%
Well-formedness judgments of \emph{linkages} have the form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
As rules \ruleref{l/empty} and \ruleref{l/add} show, a linkage is a list
of $n$~terms, each representing a field of the family modeled by the linkage.

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{n/tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

%recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$
The type~$A$ in \ruleref{l/add} and \ruleref{lsig/add},
abstracting the types of the prior fields, does not necessarily contain the same types as those recorded by~$\TyLkg{\lsig}$,
%\EDJ{Suggest changing to ``not necessarily the same as those recorded by lsig in the prior fields''},
because a field defined as the code $\wcode{\wsig}$ of a W-type has to
expose different types to different fields that come after it.
Later fields that invoke
$\Wsup{}{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate.
By contrast, $\wsig$ should be hidden
from all other fields, so that they can be reused
in a different context where the W-type signature~$\wsig$ is replaced by an extended one~$\wsig'$.

@@

\item We illustrate the structure of W-type signature (as a list) here. $\wsig_{\texttt{tm}}^0, \wsig_{\texttt{tm}}^1$ are (the subparts of) the signatures of constructors \{"tm\_unit"\} and \{"tm\_unit","tm\_app"\} respectively. Each item in this W-type signature, having two types, corresponds to one constructor, where the first type stores the non-recursive part of the constructor and the second type stores the ``arity'' of the  recursive part. For example, the type of $"app" : ("tm" \to "tm" \to "tm")$ $\cong (\bbone \times (\bbtwo \to "tm") \to "tm")$, corresponds to the premise of rule \ruleref{tm/wsup} (where $\wsigproj{0}{1}{\wsig_{\texttt{tm}}^1} = \bbone, \wsigproj{0}{2}{\wsig_{\texttt{tm}}^1} = \bbtwo$). This type means there is no information on non-recursive part (thus use unit type $\bbone$) and the ``arity'' of the recursive part is $\bbtwo$ and thus $\wsig_{\texttt{tm}}^1 = \WSigAdd{\wsig_{\texttt{tm}}^0}{\bbone}{\bbtwo}$.  The story is similar for $\wsig_{\texttt{tm}}^0$, where there is no recursive part at all and thus the ``arity'' is zero.

Generally speaking, the expressiveness of W-types comes from the ``arity'', as it encodes the case when recursive structure stores as the output of a function, mirroring the strict positivity of Inductive type.
\item 
The next four fields model the four constructors of~\lsti{tm}.
Constructor \lsti{tm_unit} is modeled as $\Wsup{}{\wsig_{\texttt{tm}}}{\top}{\bot}$
and typed with $\El{\mathit{self}.\texttt{tm}}$, where $\mathit{self}$ stands for
the typing context containing the first field~\lsti{tm}.
The W-type signature $\wsig_{\texttt{tm}}$ is exposed in
this typing context; that is, \lsti{tm} has type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$.
So $\El{\mathit{self}.\texttt{tm}}$ and $\El{\wcode{\wsig_{\texttt{tm}}}}$ can be equated,
as required by rule \ruleref{tm/wsup}.

@@

The third premise
$\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$
%\EDJ{$\var{0}$ is not identity (substitution), but the first variable, right? Or you meant to say it is identity ``sealing''?}\YZreply{Yes, I meant identity sealing. I cannot say 'sealing' because the paper currently does not use the word sealing anywhere.}\EDJreply{What about just remove the parenthesis and ``the identity''}
and $A$ is $\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are $\var{0}$,
and $\seal_6$ is from
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
to
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

@@

Inheritance and code reuse can already be expressed through the projection
of fields out of linkages and their inclusion into new linkages.
To make common patterns of linkage manipulations more convenient,
\TT provides a ``library'' of \emph{linkage transformers}, whose well-formedness judgments
have form $\goodInh{\Gm}{h}{\lsig_1}{\lsig_2}$. The idea is that
applying~$h$ to a linkage of type~$\TyLkg{\lsig_1}$ yields a linkage of
type~$\TyLkg{\lsig_2}$ (\ruleref{tm/inh}).
%Linkage transformers are second-class; the only way to use them is to
%apply them to linkages to form new linkages.

Derived families can be modeled as linkage transformers inductively
constructed from the introduction forms
$\small\InhId$, $\small\InhExt{h}{x.t}$, $\small\InhOv{h}{x.t}$, $\small\InhInh{h}$, etc.
\cref{fig:fmltt-selected} shows the $\beta$-rule of an exemplary transformer,
\ruleref{tmeq/ov/beta}.
It states that applying the transformer $\small\InhOv{h}{\mathit{self}_2.t_2}$ to
a linkage of form $\LkgAdd{\lkg}{\mathit{self}_1.t_1}$
overrides the linkage's last field~$t_1$ with~$t_2$.
For instance, the construction below shows that 
$\small\InhOv{\InhId}{\wcode{\wsig'_{\texttt{tm}}}}$
is used as the first step in creating a linkage transformer modeling a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$:
\vspace{.9ex}

@@

% The definitive version containing all the rules in \TT is stated in
% terms of a meta\-language supporting quotient inductive-inductive types
% (QIITs)~\cite{nordvall2010inductive,altkap2016}.
% QIITs support equality constructors, which facilitate the expression of
% conversion rules (e.g., \ruleref{tmeq/sub/id}).
% The formalization is available as a supplemental file in Agda syntax,
% though it is not checked by Agda because Agda does not support QIITs
% natively.
% An effort to mechanize the formalization is underway, using Agda's
% \lsti{REWRITE} pragma to replace the limited uses of equality reflection.

>>>>

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT.
Together with the identity type $\TyId{\cDot}{\cDot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
Our enhanced W-type is decided by a W-type signature $\goodWSig{\Gm}{\wsig}{n}$ 
composed of a list (\ruleref{n/wsig/empty} and \ruleref{n/wsig/add}) of $n$ pairs of 
types $\{(\wsigproj{i}{1}{\wsig}, \wsigproj{i}{2}{\wsig})\}_i$,
modeling an inductive type of $n$ constructors. 
%
Rule~\ruleref{n/tm/wsup} states that the $i$-th pair $\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and 
$\goodType{\Gm,\wsigproj{i}{1}{\wsig}}{\wsigproj{i}{2}{\wsig}}{}$ models the $i$-th constructor---given 
two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm, \mathit{self} : \nsub{\wsigproj{i}{2}{\wsig}}{x}{t_1}}{t_2}{\El{\wcode{\wsig}}}$,
we can construct a term $\Wsup{i}{\wsig}{t_1}{\mathit{self}.t_2}$ of the W-type. $\wcode{\wsig}$ gives the code of the
W-type (\ruleref{n/tm/w}).

W-types are eliminated with the form $\Wrec{\wsig}{\lkg}{t}$, where $t$ is of a
W-type $\El{\wcode{\wsig}}$, and $\lkg$ is essentially an $n$-tuple of
case handlers for the $n$ constructors in~$\wsig$ (\ruleref{n/tm/wrec}).
Each case handler has a type of form $\CaseSig{A}{B}{T}$, where
$T$ is the motive of the recursion (\ruleref{n/tyeq/casety});
for simplicity, we model non-dependent motives for recursion.
The collection of case handlers~$\lkg$ encodes those defined and inherited
by an \lsti{FRecursion} command in our plugin.
We choose to type it with a linkage type $\TyLkg{\RecSig{\wsig}{T}}$
to avoid introducing non-dependent $n$-tuples, which linkages generalize.

@@

Rules \ruleref{n/tm/wsup} and \ruleref{n/tm/wrec} require access to~$\wsig$:
the W-type is exhaustively generated by its constructors, and
its elimination must exhaustively handle all the constructors in
its signature.

@@

Well-formedness judgments of \emph{linkage signatures} have the form
$\goodSig{\Gm}{\lsig}{n}$.
As rules \ruleref{n/lsig/empty} and \ruleref{n/lsig/add} show, a linkage
signature is a list of $n$~types.
%
Well-formedness judgments of \emph{linkages} have the form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
As rules \ruleref{n/l/empty} and \ruleref{n/l/add} show, a linkage is a list
of $n$~terms, each representing a field of the family modeled by the linkage.

In rule \ruleref{n/l/add}, the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{n/tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{n/l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{n/tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

%recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$
The type~$A$ in \ruleref{n/l/add} and \ruleref{n/lsig/add},
abstracting the types of the prior fields, does not necessarily contain the same types as those recorded by~$\TyLkg{\lsig}$,
%\EDJ{Suggest changing to ``not necessarily the same as those recorded by lsig in the prior fields''},
because a field defined as the code $\wcode{\wsig}$ of a W-type has to
expose different types to different fields that come after it.
Later fields that invoke
$\Wsup{}{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as rules \ruleref{n/tm/wsup}
and \ruleref{n/tm/wrec} stipulate.
By contrast, $\wsig$ should be hidden
from all other fields, so that they can be reused
in a different context where the W-type signature~$\wsig$ is replaced by an extended one~$\wsig'$.

@@

\item We illustrate the structure of W-type signature (as a list) here. $\wsig_{\texttt{tm}}^0, \wsig_{\texttt{tm}}^1$ are (the subparts of) the signatures of constructors \{"tm\_unit"\} and \{"tm\_unit","tm\_app"\} respectively. Each item in this W-type signature, having two types, corresponds to one constructor, where the first type stores the non-recursive part of the constructor and the second type stores the ``arity'' of the  recursive part. For example, the type of $"app" : ("tm" \to "tm" \to "tm")$ $\cong (\bbone \times (\bbtwo \to "tm") \to "tm")$, corresponds to the premise of rule \ruleref{n/tm/wsup} (where $\wsigproj{0}{1}{\wsig_{\texttt{tm}}^1} = \bbone, \wsigproj{0}{2}{\wsig_{\texttt{tm}}^1} = \bbtwo$). This type means there is no information on non-recursive part (thus use unit type $\bbone$) and the ``arity'' of the recursive part is $\bbtwo$ and thus $\wsig_{\texttt{tm}}^1 = \WSigAdd{\wsig_{\texttt{tm}}^0}{\bbone}{\bbtwo}$.  The story is similar for $\wsig_{\texttt{tm}}^0$, where there is no recursive part at all and thus the ``arity'' is zero.

Generally speaking, the expressiveness of W-types comes from the ``arity'', as it encodes the case when recursive structure stores as the output of a function, mirroring the strict positivity of Inductive type.
\item 
The next four fields model the four constructors of~\lsti{tm}.
Constructor \lsti{tm_unit} is modeled as $\Wsup{}{\wsig_{\texttt{tm}}}{\top}{\bot}$
and typed with $\El{\mathit{self}.\texttt{tm}}$, where $\mathit{self}$ stands for
the typing context containing the first field~\lsti{tm}.
The W-type signature $\wsig_{\texttt{tm}}$ is exposed in
this typing context; that is, \lsti{tm} has type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$.
So $\El{\mathit{self}.\texttt{tm}}$ and $\El{\wcode{\wsig_{\texttt{tm}}}}$ can be equated,
as required by rule \ruleref{n/tm/wsup}.

@@

The third premise
$\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$
in rules \ruleref{n/l/add} and \ruleref{n/lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{n/tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{n/tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$
%\EDJ{$\var{0}$ is not identity (substitution), but the first variable, right? Or you meant to say it is identity ``sealing''?}\YZreply{Yes, I meant identity sealing. I cannot say 'sealing' because the paper currently does not use the word sealing anywhere.}\EDJreply{What about just remove the parenthesis and ``the identity''}
and $A$ is $\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are $\var{0}$,
and $\seal_6$ is from
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
to
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

@@

Inheritance and code reuse can already be expressed through the projection
of fields out of linkages and their inclusion into new linkages.
To make common patterns of linkage manipulations more convenient,
\TT provides a ``library'' of \emph{linkage transformers}, whose well-formedness judgments
have form $\goodInh{\Gm}{h}{\lsig_1}{\lsig_2}$. The idea is that
applying~$h$ to a linkage of type~$\TyLkg{\lsig_1}$ yields a linkage of
type~$\TyLkg{\lsig_2}$ (\ruleref{n/tm/inh}).
%Linkage transformers are second-class; the only way to use them is to
%apply them to linkages to form new linkages.

Derived families can be modeled as linkage transformers inductively
constructed from the introduction forms
$\small\InhId$, $\small\InhExt{h}{x.t}$, $\small\InhOv{h}{x.t}$, $\small\InhInh{h}$, etc.
\cref{fig:fmltt-selected} shows the $\beta$-rule of an exemplary transformer,
\ruleref{n/tmeq/ov/beta}.
It states that applying the transformer $\small\InhOv{h}{\mathit{self}_2.t_2}$ to
a linkage of form $\LkgAdd{\lkg}{\mathit{self}_1.t_1}$
overrides the linkage's last field~$t_1$ with~$t_2$.
For instance, the construction below shows that 
$\small\InhOv{\InhId}{\wcode{\wsig'_{\texttt{tm}}}}$
is used as the first step in creating a linkage transformer modeling a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$:
\vspace{.9ex}

@@

% The definitive version containing all the rules in \TT is stated in
% terms of a meta\-language supporting quotient inductive-inductive types
% (QIITs)~\cite{nordvall2010inductive,altkap2016}.
% QIITs support equality constructors, which facilitate the expression of
% conversion rules (e.g., \ruleref{n/tmeq/sub/id}).
% The formalization is available as a supplemental file in Agda syntax,
% though it is not checked by Agda because Agda does not support QIITs
% natively.
% An effort to mechanize the formalization is underway, using Agda's
% \lsti{REWRITE} pragma to replace the limited uses of equality reflection.
